The Talking Heads Experiment

### 2015-04-16 chap7/chap7.tex
minor changes
hyphenation, orphans, widows and some other layout updated

But some drastic events happened in Wittgenstein's life which 
caused a radical shift in his philosophical orientation. 
One of them was that he became headmaster of a small
school in the Austrian mountains,
without losing his interest for philosophising about language and 
meaning. By working with children and by seeing concretely
how they engaged with language, Wittgenstein realised that the 
logicist approach did not address some basic questions one 
can ask about language, particularly how the semantics of a 
language might arise, or how people ever develop a shared 
communication system. Wittgenstein
saw clearly that a language imposes a certain view on the 
world, which gives each language its own strength as well as
its limitations. The logicist 
framework might be ideal as a tool for constructing a 
{\itshape post factum} formal description of the semantics of
an (ideal) language, just as a structuralist grammar identifies
the frozen state of its syntax, but
not for modeling how language achieves its communicative purpose, 
comes about, or evolves.

Wittgenstein proposed to study the use of language in terms of 
games. This notion captures that the most basic form of
language involves a social interaction between individuals
within a specific setting, which acts as a context restricting
the possible meanings. The interaction is played out 
along a conventionalised set of rules, as is chess or any
other game. The notion of a game captures
that a language interaction has a purpose, for example,
to identify an object in a context, to gather 
information, to transmit emotion, to try and
invoke an action by another person, etc. Words get 
their meaning as part of a language game. Language is a tool 
fully integrated in the rest of human social activity.\footnote{See in particular \cite{Wittgenstein:1953}.}

@@

I now show a series of example games taken from 
an experiment in which two situated embodied agents,
{\bfshape a1} and {\bfshape a2}, play grounded language games based on 
coloured figures pasted on the white board in front
of them. The following channels are available to the agents 
in the experiments in this section: 
HPOS (the horizontal position of the midpoint 
of the segment's bounding box), 
VPOS (the vertical position of the midpoint of the segment's 
bounding box), HEIGHT (the height of the bounding box), 
WIDTH (the width of the bounding box), AREA
(the area of the segment, calculated
by counting the number of pixels that belong to it), 
R (the average redness of the pixels in the segment), 
G (the average greenness of the pixels in the segment), 
B (the average blueness of the pixels in the segment). 
These ``colour'' channels are not to be confused with the 
human opponent colour channels so the distinctions 
that form of them are not directly perceivable by 
human observers.

@@

\begin{table}
\begin{center}
\begin{tabular}{ l  l  l }
\lsptoprule
{\itshape channel}& {\itshape obj-0} & {\itshape obj-1}\\ \midrule
HPOS & 0.27 & 0.16\\ 
VPOS & 0.20 & 0.20\\ 
HEIGHT & 0.15 & 0.15\\ 
WIDTH & 0.10 & 0.11\\ 
AREA & 0.10 & 0.10\\ 
R & 0.23 & 0.25\\ 
G & 0.32 & 0.34\\ 
B & 0.63 & 0.65\\ 
\lspbottomrule
\end{tabular}
\caption{\label{tab:game3b}First words after about a dozen games.}
\end{center}
\end{table}

HPOS is the most salient channel. After sensor-scaling, 
the two values for HPOS are still be drawn further
apart with 1.0 for object-0 and 
0.0 for object-1 so that the category [HPOS 0.5–1.0] easily 
distinguishes the topic (object-0) from object-1.

The left image is that of the hearer {\bfshape a1}, the right one
that of the speaker, {\bfshape a2}.
{\bfshape a2} had already invented the word `gofubo' for 
[HPOS 0.5–1.0] (to the right) in an earlier game, 
but the word was not acquired in that game
by {\bfshape a1} because he still missed the 
appropriate distinction. Meanwhile the HPOS category 
is available to {\bfshape a1} due to an
expansion of his discrimination networks
and so he can store the word: 
\begin{verbatim}
Game 3 
  a2 is the speaker. a1 is the hearer. 
  a2 segments the context into 2 objects: 
       object-0 object-1
  a2 chooses object-0 as the topic 
  a2 categorises the topic as [HPOS 0.5–1.0]
  a2 says: gofubo
  a1 does not know gofubo
  a1 says: gofubo?
  a2 points to object-0
  a1 categorises the topic as [HPOS 0.5–1.0]
  a1 stores gofubo as [HPOS 0.5–1.0]
\end{verbatim}
This game proceeds essentially like in the computer
simulations studied before. The major difference is
that now real images have been used, the objects
considered are the outcome of segmentation processes,
the sensory characteristics have been derived from the 
image itself, and the pointing has been done by 
physically moving the cameras.

@@

Immediately the agents apply this word to very 
different scenes, such as the ones in \figref{fig:plate-10} (middle and bottom). The middle picture
shows on the left the segmented scene
from the speaker in game 5, and on the right the one
from the hearer. In game 5, the figures are blue
rectangles. They are much further apart than 
the two circles in game 3. Nevertheless, 
after scaling, they are categorised and
conceptualised similarly and therefore
the same word could be used effectively. The concept 
of [HPOS 0.5–1.0] and hence the word `gofubo' 
is general {\itshape from the very start}. 
The agents do not need to see many examples because they do 
not use inductive generalisation. Instead, 
the HPOS category is constructed in a top-down fashion 
as soon as the HPOS channel has been salient
and is immediately available for use in the 
discrimination game and hence in verbalising 
the scene. This explains why the word-meaning acquisition
process observed in the experiments goes so 
amazingly fast.

The bottom of \figref{fig:plate-10}
shows yet another scene where the word `gofubo' 
was used with success. The topic is the rightmost
shape in the scene and so [HPOS 0.5–1.0] is once more 
distinctive. To an outside observer it may look 
like the agents have performed a gigantic inductive
leap, but this is not the case at all. The agents
do not try to abstract the commonalities from 
different examples but construct distinctions in a top 
down fashion and try to apply them to the perceived image.

@@

\begin{table}
\begin{center}
\begin{tabular}{ l  l  l  l  l }
\lsptoprule
{\itshape Word}&{\itshape Meaning}&{\itshape Translation} & {\bfshape a1}&{\bfshape a2} \\ \midrule
wawosido & [HPOS 0.0–0.5] &left&0.4&0.4\\ 
meluri & [HPOS 0.25–0.5] &medium left&0.1&0.0\\ 
gofubo & [HPOS 0.5–1.0]& right&1.0&1.0\\ 
wiwigapo & [VPOS 0.0–0.5] &left&0.1&0.0\\ 
fozumoba & [AREA 0.5–1.0]&large & 0.0&0.1\\ 
wefoto & [R 0.0–0.5]& low redness &0.2&0.2\\ 
togene & [R 0.5–1.0]& high redness &0.5&1.0\\ 
fumudanu & [G 0.0–0.5]& low greenness &0.2&0.2\\ 
puxedu & [G 0.5–1.0]& high greenness &0.4&0.4\\ 
\lspbottomrule
\end{tabular}
\caption{\label{tab:game50}First words after about a dozen games.}
\end{center}
\end{table}

@@

\begin{table}
\begin{center}
\begin{tabular}{ l  l  l  l  l }
\lsptoprule
{\itshape Word}&{\itshape Meaning}&{\itshape Translation} & {\bfshape a1}&{\bfshape a2} \\ \midrule
wawosido & [HPOS 0.0–0.5] &left&0.7&0.5\\ 
meluri & [HPOS 0.25–0.5] &medium left&0.4&0.4\\ 
gofubo & [HPOS 0.5–1.0]& right&1.0&1.0\\ 
vokomutu & [HPOS 0.5–0.75] &medium right&0.2&0.2\\ 
buwonipo & [HPOS 0.75–0.875] &strongly right&0.1&0.0\\ 
wiwigapo & [VPOS 0.0–0.5] &down&0.6&0.6\\ 
fozumoba & [AREA 0.5–1.0]&large & 0.2&0.4\\ 
wefoto & [R 0.0–0.5]& low redness &0.2&0.3\\ 
togene & [R 0.5–1.0]& high redness &1.0&1.0\\ 
fumudanu & [G 0.0–0.5]& low greenness &0.7&0.7\\ 
puxedu & [G 0.5–1.0]& high greenness &0.7&0.9\\ 
\lspbottomrule
\end{tabular}
\caption{\label{tab:goubo}Population lexicon after 100 games.}
\end{center}
\end{table}

@@

One thing is striking about this lexicon. It contains\is{environmental influence}
words for fine-grained distinctions along the 
horizontal position axis, but none for width or 
height. Indeed if we look at the discrimination
trees as they exist at this point (figure 
\ref{discri200} top) we see that there are no 
discrimination trees for the HEIGHT and WIDTH
channels. This is entirely due to the 
environment. There have simply been no situations on 
the white board yet where these distinctions are salient 
enough.

@@

\begin{figure}
\begin{center}
\includegraphics[width=0.8\columnwidth]{chap7/figs/plate-13.pdf}
\end{center}
\caption{Top and bottom: Examples of new segmented scenes that stimulate conceptual development
and hence expansions of the lexicon. The most salient characteristic in the top scene is WIDTH and in the bottom 
scene HEIGHT.}
\label{fig:plate-13}
\end{figure}

@@

\begin{table}
\begin{center}
\begin{tabular}{ l  l  l }
\lsptoprule
{\itshape channel}& {\itshape obj-0} & {\itshape obj-1}\\ \midrule
HPOS & 0.08 & 0.09\\ 
VPOS & 0.21 & 0.10\\ 
HEIGHT & 0.10 & 0.15\\ 
WIDTH & 0.42 & 0.11\\ 
AREA & 0.34 & 0.16\\ 
R & 0.33 & 0.31\\ 
G & 0.68 & 0.65\\ 
B & 0.52 & 0.50\\ 
\lspbottomrule
\end{tabular}
\caption{\label{tab:game126}Sensory data for game 126.}
\end{center}
\end{table}
Clearly width is the most salient channel for the speaker. 
It is therefore chosen and because the 
agents have already grown distinctions on this channel, 
discrimination succeeds and a new word can be
constructed and stored. 
\begin{verbatim}
Game 126 
  a2 is the speaker. a1 is the hearer. 
  a2 segments the context into 2 objects: 
       object-0 object-1
  a2 chooses object-0 as the topic 
  a2 categorises the topic as [WIDTH 0.5–1.0]
  a2 creates a new word: vaviwumu
  a2 says: vaviwumu
  a1 does not know vaviwumu
  a1 says: vaviwumu?
  a2 points to object-0
  a1 categorises the topic as [WIDTH 0.5–1.0]
  a1 stores vaviwumu as [WIDTH 0.5–1.0]
\end{verbatim}
\figref{fig:plate-13} (bottom) contains another example
of segmented images which have stimulated conceptual
growth and hence an expansion of the lexicon. 
In this case, categorisations on the HEIGHT channel are
relevant and a word developed for tall.

\begin{figure}
\begin{center}
\includegraphics[width=0.8\columnwidth]{chap7/figs/plate-14-area.pdf}
\end{center}
\caption{Segmented images where distinctions on the AREA channel have been used. The topic (with dashed bounding 
box) in the two top cases has been categorised as large. In the bottom case, a word meaning `small' was used.}
\label{fig:plate-14}
\end{figure}

\figref{fig:plate-14} shows another series of image segments, 
where categorisations based on the AREA channel were effective
and \figref{fig:plate-15} shows additional image segments exercising
VPOS, HPOS and colour distinctions.

\begin{figure}
\begin{center}
\includegraphics[width=0.8\columnwidth]{chap7/figs/plate-15-vpos.pdf}
\end{center}
\caption{The image segments from the top game require 
distinctions on the VPOS channel and 
caused the creation of a word for `upper'. The image 
segments in the middle lead to the use of refined
distinctions on the HPOS channel, so as to 
identify the middle square. The topic in the 
image segments at the bottom was done with a 
conjunction of categories: blue and tall.}
\label{fig:plate-15}
\end{figure}

@@

\begin{table}
\begin{center}
\begin{tabular}{ l  l  l  l  l }
\lsptoprule
{\itshape Word}&{\itshape Meaning}&{\itshape Translation} & {\bfshape a1}&{\bfshape a2} \\ \midrule
wawosido & [HPOS 0.0–0.5] &left&0.9&0.4\\ 
gixepo & [HPOS 0.0–0.25] & very left&0.4&0.4\\ 
wonuxa & [HPOS 0.0–0.25] & very left&0.10&0.0\\ 
meluri & [HPOS 0.25–0.5] &medium left&0.4&0.4\\ 
gofubo & [HPOS 0.5–1.0]& right&1.0&1.0\\ 
vokomutu & [HPOS 0.5–0.75] &medium right&0.4&0.4\\ 
buwonipo & [HPOS 0.75–0.875] &strongly right&0.1&0.0\\ 
wiwigapo & [VPOS 0.0–0.5] &down&0.5&0.4\\ 
putuwenu & [VPOS 0.5–1.0]&up & 0.5&0.5\\ 
vaviwumu & [WIDTH 0.5–1.0]&wide & 0.5&0.5\\ 
pesidumu & [AREA 0.0–0.5]&small& 0.2&0.2\\ 
fozumoba & [AREA 0.5–1.0]&large & 1.0&1.0\\ 
wefoto & [R 0.0–0.5]& low redness &0.2&0.1\\ 
togene & [R 0.5–1.0]& high redness &0.9&1.0\\ 
fumudanu & [G 0.0–0.5]& low greenness &0.7&0.9\\ 
puxedu & [G 0.5–1.0]& high greenness &0.9&0.9\\ 
\lspbottomrule
\end{tabular}
\caption{\label{tab:upper}Population lexicon after 200 games.}
\end{center}
\end{table}
Note that new words have come into the lexicon 
for up (`putuwenu'), down (`wiwigapo'), and 
wide (`vaviwumu'). The discrimination trees of 
the WIDTH and HEIGHT channel have started to 
expand (\figref{discri200} bottom).

@@

Grounding language games in physical environments\is{perceptual anomalies}
makes it obviously harder
for the agents in a number of respects. First of 
all because perception and segmentation may differ, 
the salient characteristics of objects may not be the same
and as a result the hearer may guess another meaning for an 
unknown word, compared to 
the one used by the speaker. This happens for
example in the following game (game 128) which is based on 
the segmented images shown in \figref{fig:plate-16} (top). The
speaker's perception is shown to the left and the 
hearer's to the right. 
\begin{verbatim}
Game 128
  a1 is the speaker. a2 is the hearer. 
  a1 segments the context into 3 objects: 
       object-0 object-1 object-2
  a1 chooses object-0 as the topic 
  a1 categorises the topic as 
       [HPOS 0.0–0.25] [HPOS 0.0–0.125]
  a1 creates a new word: `bagaxe' for [HPOS 0.0–0.25]
  a1 says: bagaxe
  a2 does not know bagaxe
  a2 says: bagaxe?
  a1 points to object-0
  a2 categorises the topic as [AREA 0.5–1.0]
  a2 stores bagaxe as [AREA 0.5–1.0]
\end{verbatim}

@@

\begin{table}
\begin{center}
\begin{tabular}{ l  l  l  l }
\lsptoprule
{\itshape channel}& {\itshape obj-0} & {\itshape obj-1} & {\itshape obj-2}\\ 
HPOS & 0.03 & 0.15 & 0.33\\ 
VPOS & 0.29 & 0.29 & 0.01\\ 
HEIGHT & 0.37 & 0.39 & 0.07\\ 
WIDTH & 0.10 & 0.08 & 0.27\\ 
AREA & 0.29 & 0.22 & 0.16\\ 
R & 0.95 & 0.97 & 0.97 \\ 
G & 0.33 & 0.40 & 0.93\\ 
B & 0.35 & 0.44 & 0.26\\ 
\lspbottomrule
\end{tabular}
\caption{\label{tab:game128}Sensory data for game 128.}
\end{center}
\end{table}
Context-scaling further
amplifies this difference which shows that it is not 
always beneficial to do so. {\bfshape a1} uses
the horizontal axis, creating a new word `bagaxe' 
meaning `very left' [HPOS 0.0–0.25] whereas {\bfshape a2}, for
whom the area is the most salient characteristic, 
associates this word with [AREA 0.5–1.0] (large). 
This example shows that grounding introduces 
additional risks for the introduction of semantic
incoherence in a group's lexicon.

>>>>

But some drastic events happened in Wittgenstein's life which 
caused a radical shift in his philosophical orientation. 
One of them was that he became headmaster of a small
school in the Austrian mountains,
without losing his interest for philosophising about language and 
meaning. By working with children and by seeing concretely
how they engaged with language, Wittgenstein realised that the 
logicist approach did not address some basic questions one 
can ask about language, particularly how the semantics of a 
language might arise, or how people ever develop a shared 
communication system. Wittgenstein
saw clearly that a language imposes a certain view on the 
world, which gives each language its own strength as well as
its limitations. The logicist 
framework might be ideal as a tool for constructing a 
{\itshape post factum} formal description of the semantics of
an (ideal) language, just as a structuralist grammar identifies
the frozen state of its syntax, but
not for modeling how language achieves its communicative purpose, 
comes about, or evolves. 
\enlargethispage{1\baselineskip}
Wittgenstein proposed to study the use of language in terms of 
games. This notion captures that the most basic form of
language involves a social interaction between individuals
within a specific setting, which acts as a context restricting
the possible meanings. The interaction is played out 
along a conventionalised set of rules, as is chess or any
other game. The notion of a game captures
that a language interaction has a purpose, for example,
to identify an object in a context, to gather 
information, to transmit emotion, to try and
invoke an action by another person, etc. Words get 
their meaning as part of a language game. Language is a tool 
fully integrated in the rest of human social activity.\footnote{See in particular \cite{Wittgenstein:1953}.}

@@

I now show a series of example games taken from 
an experiment in which two situated embodied agents,
{\bfshape a1} and {\bfshape a2}, play grounded language games based on 
coloured figures pasted on the white board in front
of them. The following channels are available to the agents 
in the experiments in this section: 
\textsc{hpos} (the horizontal position of the midpoint 
of the segment's bounding box), 
\textsc{vpos} (the vertical position of the midpoint of the segment's 
bounding box), \textsc{height} (the height of the bounding box), 
\textsc{width} (the width of the bounding box), \textsc{area}
(the area of the segment, calculated
by counting the number of pixels that belong to it), 
R (the average redness of the pixels in the segment), 
G (the average greenness of the pixels in the segment), 
B (the average blueness of the pixels in the segment). 
These ``colour'' channels are not to be confused with the 
human opponent colour channels so the distinctions 
that form of them are not directly perceivable by 
human observers.

@@

\begin{table}[b]
\begin{center}
\begin{tabular}{lll}
\lsptoprule
{\itshape channel}& {\itshape obj-0} & {\itshape obj-1}\\ \midrule
\textsc{hpos} & 0.27 & 0.16\\ 
\textsc{vpos} & 0.20 & 0.20\\ 
\textsc{height} & 0.15 & 0.15\\ 
\textsc{width} & 0.10 & 0.11\\ 
\textsc{area} & 0.10 & 0.10\\ 
R & 0.23 & 0.25\\ 
G & 0.32 & 0.34\\ 
B & 0.63 & 0.65\\ 
\lspbottomrule
\end{tabular}
\caption{\label{tab:game3b}First words after about a dozen games.}
\end{center}
\end{table}
\clearpage
\textsc{hpos} is the most salient channel. After sensor-scaling, 
the two values for \textsc{hpos} are still be drawn further
apart with 1.0 for object-0 and 
0.0 for object-1 so that the category [\textsc{hpos} 0.5–1.0] easily 
distinguishes the topic (object-0) from object-1.

The left image is that of the hearer {\bfshape a1}, the right one
that of the speaker, {\bfshape a2}.
{\bfshape a2} had already invented the word `gofubo' for 
[\textsc{hpos} 0.5–1.0] (to the right) in an earlier game, 
but the word was not acquired in that game
by {\bfshape a1} because he still missed the 
appropriate distinction. Meanwhile the \textsc{hpos} category 
is available to {\bfshape a1} due to an
expansion of his discrimination networks
and so he can store the word: 
\begin{verbatim}
Game 3 
  a2 is the speaker. a1 is the hearer. 
  a2 segments the context into 2 objects: 
       object-0 object-1
  a2 chooses object-0 as the topic 
  a2 categorises the topic as [\textsc{hpos} 0.5–1.0]
  a2 says: gofubo
  a1 does not know gofubo
  a1 says: gofubo?
  a2 points to object-0
  a1 categorises the topic as [\textsc{hpos} 0.5–1.0]
  a1 stores gofubo as [\textsc{hpos} 0.5–1.0]
\end{verbatim}
This game proceeds essentially like in the computer
simulations studied before. The major difference is
that now real images have been used, the objects
considered are the outcome of segmentation processes,
the sensory characteristics have been derived from the 
image itself, and the pointing has been done by 
physically moving the cameras.

@@

Immediately the agents apply this word to very 
different scenes, such as the ones in \figref{fig:plate-10} (middle and bottom). The middle picture
shows on the left the segmented scene
from the speaker in game 5, and on the right the one
from the hearer. In game 5, the figures are blue
rectangles. They are much further apart than 
the two circles in game 3. Nevertheless, 
after scaling, they are categorised and
conceptualised similarly and therefore
the same word could be used effectively. The concept 
of [\textsc{hpos} 0.5–1.0] and hence the word `gofubo' 
is general {\itshape from the very start}. 
The agents do not need to see many examples because they do 
not use inductive generalisation. Instead, 
the \textsc{hpos} category is constructed in a top-down fashion 
as soon as the \textsc{hpos} channel has been salient
and is immediately available for use in the 
discrimination game and hence in verbalising 
the scene. This explains why the word-meaning acquisition
process observed in the experiments goes so 
amazingly fast.

The bottom of \figref{fig:plate-10}
shows yet another scene where the word `gofubo' 
was used with success. The topic is the rightmost
shape in the scene and so [\textsc{hpos} 0.5–1.0] is once more 
distinctive. To an outside observer it may look 
like the agents have performed a gigantic inductive
leap, but this is not the case at all. The agents
do not try to abstract the commonalities from 
different examples but construct distinctions in a top 
down fashion and try to apply them to the perceived image.

@@

\begin{table}
\begin{center}
\begin{tabular}{lllll}
\lsptoprule
{\itshape Word}&{\itshape Meaning}&{\itshape Translation} & {\bfshape a1}&{\bfshape a2} \\ \midrule
wawosido & [\textsc{hpos} 0.0–0.5] &left&0.4&0.4\\ 
meluri & [\textsc{hpos} 0.25–0.5] &medium left&0.1&0.0\\ 
gofubo & [\textsc{hpos} 0.5–1.0]& right&1.0&1.0\\ 
wiwigapo & [\textsc{vpos} 0.0–0.5] &left&0.1&0.0\\ 
fozumoba & [\textsc{area} 0.5–1.0]&large & 0.0&0.1\\ 
wefoto & [R 0.0–0.5]& low redness &0.2&0.2\\ 
togene & [R 0.5–1.0]& high redness &0.5&1.0\\ 
fumudanu & [G 0.0–0.5]& low greenness &0.2&0.2\\ 
puxedu & [G 0.5–1.0]& high greenness &0.4&0.4\\ 
\lspbottomrule
\end{tabular}
\caption{\label{tab:game50}First words after about a dozen games.}
\end{center}
\end{table}

@@

\begin{table}
\begin{center}
\begin{tabular}{lllll}
\lsptoprule
{\itshape Word}&{\itshape Meaning}&{\itshape Translation} & {\bfshape a1}&{\bfshape a2} \\ \midrule
wawosido & [\textsc{hpos} 0.0–0.5] &left&0.7&0.5\\ 
meluri & [\textsc{hpos} 0.25–0.5] &medium left&0.4&0.4\\ 
gofubo & [\textsc{hpos} 0.5–1.0]& right&1.0&1.0\\ 
vokomutu & [\textsc{hpos} 0.5–0.75] &medium right&0.2&0.2\\ 
buwonipo & [\textsc{hpos} 0.75–0.875] &strongly right&0.1&0.0\\ 
wiwigapo & [\textsc{vpos} 0.0–0.5] &down&0.6&0.6\\ 
fozumoba & [\textsc{area} 0.5–1.0]&large & 0.2&0.4\\ 
wefoto & [R 0.0–0.5]& low redness &0.2&0.3\\ 
togene & [R 0.5–1.0]& high redness &1.0&1.0\\ 
fumudanu & [G 0.0–0.5]& low greenness &0.7&0.7\\ 
puxedu & [G 0.5–1.0]& high greenness &0.7&0.9\\ 
\lspbottomrule
\end{tabular}
\caption{\label{tab:goubo}Population lexicon after 100 games.}
\end{center}
\end{table}

@@

One thing is striking about this lexicon. It contains\is{environmental influence}
words for fine-grained distinctions along the 
horizontal position axis, but none for width or 
height. Indeed if we look at the discrimination
trees as they exist at this point (figure 
\ref{discri200} top) we see that there are no 
discrimination trees for the \textsc{height} and \textsc{width}
channels. This is entirely due to the 
environment. There have simply been no situations on 
the white board yet where these distinctions are salient 
enough.

@@

\begin{figure}
\begin{center}
\includegraphics[width=0.8\columnwidth]{chap7/figs/plate-13.pdf}
\end{center}
\caption{Top and bottom: Examples of new segmented scenes that stimulate conceptual development
and hence expansions of the lexicon. The most salient characteristic in the top scene is \textsc{width} and in the bottom 
scene \textsc{height}.}
\label{fig:plate-13}
\end{figure}

@@

\begin{table}
\begin{center}
\begin{tabular}{lll}
\lsptoprule
{\itshape channel}& {\itshape obj-0} & {\itshape obj-1}\\ \midrule
\textsc{hpos} & 0.08 & 0.09\\ 
\textsc{vpos} & 0.21 & 0.10\\ 
\textsc{height} & 0.10 & 0.15\\ 
\textsc{width} & 0.42 & 0.11\\ 
\textsc{area} & 0.34 & 0.16\\ 
R & 0.33 & 0.31\\ 
G & 0.68 & 0.65\\ 
B & 0.52 & 0.50\\ 
\lspbottomrule
\end{tabular}
\caption{\label{tab:game126}Sensory data for game 126.}
\end{center}
\end{table}
Clearly width is the most salient channel for the speaker. 
It is therefore chosen and because the 
agents have already grown distinctions on this channel, 
discrimination succeeds and a new word can be
constructed and stored. 
\begin{verbatim}
Game 126 
  a2 is the speaker. a1 is the hearer. 
  a2 segments the context into 2 objects: 
       object-0 object-1
  a2 chooses object-0 as the topic 
  a2 categorises the topic as [\textsc{width} 0.5–1.0]
  a2 creates a new word: vaviwumu
  a2 says: vaviwumu
  a1 does not know vaviwumu
  a1 says: vaviwumu?
  a2 points to object-0
  a1 categorises the topic as [\textsc{width} 0.5–1.0]
  a1 stores vaviwumu as [\textsc{width} 0.5–1.0]
\end{verbatim}
\figref{fig:plate-13} (bottom) contains another example
of segmented images which have stimulated conceptual
growth and hence an expansion of the lexicon. 
In this case, categorisations on the \textsc{height} channel are
relevant and a word developed for tall.

\begin{figure}
\begin{center}
\includegraphics[width=0.8\columnwidth]{chap7/figs/plate-14-area.pdf}
\end{center}
\caption{Segmented images where distinctions on the \textsc{area} channel have been used. The topic (with dashed bounding 
box) in the two top cases has been categorised as large. In the bottom case, a word meaning `small' was used.}
\label{fig:plate-14}
\end{figure}

\figref{fig:plate-14} shows another series of image segments, 
where categorisations based on the \textsc{area} channel were effective
and \figref{fig:plate-15} shows additional image segments exercising
\textsc{vpos}, \textsc{hpos} and colour distinctions.

\begin{figure}
\begin{center}
\includegraphics[width=0.8\columnwidth]{chap7/figs/plate-15-vpos.pdf}
\end{center}
\caption{The image segments from the top game require 
distinctions on the \textsc{vpos} channel and 
caused the creation of a word for `upper'. The image 
segments in the middle lead to the use of refined
distinctions on the \textsc{hpos} channel, so as to 
identify the middle square. The topic in the 
image segments at the bottom was done with a 
conjunction of categories: blue and tall.}
\label{fig:plate-15}
\end{figure}

@@

\begin{table}
\begin{center}
\begin{tabular}{lllll}
\lsptoprule
{\itshape Word}&{\itshape Meaning}&{\itshape Translation} & {\bfshape a1}&{\bfshape a2} \\ \midrule
wawosido & [\textsc{hpos} 0.0–0.5] &left&0.9&0.4\\ 
gixepo & [\textsc{hpos} 0.0–0.25] & very left&0.4&0.4\\ 
wonuxa & [\textsc{hpos} 0.0–0.25] & very left&0.10&0.0\\ 
meluri & [\textsc{hpos} 0.25–0.5] &medium left&0.4&0.4\\ 
gofubo & [\textsc{hpos} 0.5–1.0]& right&1.0&1.0\\ 
vokomutu & [\textsc{hpos} 0.5–0.75] &medium right&0.4&0.4\\ 
buwonipo & [\textsc{hpos} 0.75–0.875] &strongly right&0.1&0.0\\ 
wiwigapo & [\textsc{vpos} 0.0–0.5] &down&0.5&0.4\\ 
putuwenu & [\textsc{vpos} 0.5–1.0]&up & 0.5&0.5\\ 
vaviwumu & [\textsc{width} 0.5–1.0]&wide & 0.5&0.5\\ 
pesidumu & [\textsc{area} 0.0–0.5]&small& 0.2&0.2\\ 
fozumoba & [\textsc{area} 0.5–1.0]&large & 1.0&1.0\\ 
wefoto & [R 0.0–0.5]& low redness &0.2&0.1\\ 
togene & [R 0.5–1.0]& high redness &0.9&1.0\\ 
fumudanu & [G 0.0–0.5]& low greenness &0.7&0.9\\ 
puxedu & [G 0.5–1.0]& high greenness &0.9&0.9\\ 
\lspbottomrule
\end{tabular}
\caption{\label{tab:upper}Population lexicon after 200 games.}
\end{center}
\end{table}
Note that new words have come into the lexicon 
for up (`putuwenu'), down (`wiwigapo'), and 
wide (`vaviwumu'). The discrimination trees of 
the \textsc{width} and \textsc{height} channel have started to 
expand (\figref{discri200} bottom).

@@

Grounding language games in physical environments\is{perceptual anomalies}
makes it obviously harder
for the agents in a number of respects. First of 
all because perception and segmentation may differ, 
the salient characteristics of objects may not be the same
and as a result the hearer may guess another meaning for an 
unknown word, compared to 
the one used by the speaker. This happens for
example in the following game (game 128) which is based on 
the segmented images shown in \figref{fig:plate-16} (top). The
speaker's perception is shown to the left and the 
hearer's to the right. 
\begin{verbatim}
Game 128
  a1 is the speaker. a2 is the hearer. 
  a1 segments the context into 3 objects: 
       object-0 object-1 object-2
  a1 chooses object-0 as the topic 
  a1 categorises the topic as 
       [\textsc{hpos} 0.0–0.25] [\textsc{hpos} 0.0–0.125]
  a1 creates a new word: `bagaxe' for [\textsc{hpos} 0.0–0.25]
  a1 says: bagaxe
  a2 does not know bagaxe
  a2 says: bagaxe?
  a1 points to object-0
  a2 categorises the topic as [\textsc{area} 0.5–1.0]
  a2 stores bagaxe as [\textsc{area} 0.5–1.0]
\end{verbatim}

@@

\begin{table}
\begin{center}
\begin{tabular}{llll}
\lsptoprule
{\itshape channel}& {\itshape obj-0} & {\itshape obj-1} & {\itshape obj-2}\\ 
\textsc{hpos} & 0.03 & 0.15 & 0.33\\ 
\textsc{vpos} & 0.29 & 0.29 & 0.01\\ 
\textsc{height} & 0.37 & 0.39 & 0.07\\ 
\textsc{width} & 0.10 & 0.08 & 0.27\\ 
\textsc{area} & 0.29 & 0.22 & 0.16\\ 
R & 0.95 & 0.97 & 0.97 \\ 
G & 0.33 & 0.40 & 0.93\\ 
B & 0.35 & 0.44 & 0.26\\ 
\lspbottomrule
\end{tabular}
\caption{\label{tab:game128}Sensory data for game 128.}
\end{center}
\end{table}
Context-scaling further
amplifies this difference which shows that it is not 
always beneficial to do so. {\bfshape a1} uses
the horizontal axis, creating a new word `bagaxe' 
meaning `very left' [\textsc{hpos} 0.0–0.25] whereas {\bfshape a2}, for
whom the area is the most salient characteristic, 
associates this word with [\textsc{area} 0.5–1.0] (large). 
This example shows that grounding introduces 
additional risks for the introduction of semantic
incoherence in a group's lexicon.

Due to perceptual anomalies, a game may fail even though 
both agents already have the same meaning for the 
same word. This happens when this meaning yields different 
objects (or no objects at all) for the speaker and the hearer. 
As a consequence, the hearer adopts another meaning for 
the word which starts to compete with the one he already had. 
This happened in the following game (game 127)
based on the segmented images shown in \figref{fig:plate-16} (bottom). The speaker (left image) has identified
only two objects, but the hearer three. The 
speaker (left image) uses green as distinguishing 
characteristic which is indeed appropriate. but 
because the hearer's third object (the top rectangle) is 
also green, this distinction fails for him.  
Even though the hearer had already a well established
meaning for `puxedu', he associates a new meaning 
based on the \textsc{width}-channel with this word because this 
channel is now the most salient. 
\begin{verbatim}
Game 127
  a1 is the speaker. a2 is the hearer. 
  a1 segments the context into 2 objects: 
       s-object-0 s-object-1
  a1 chooses object-0 as the topic 
  a1 considers as salient G R \textsc{area} \textsc{hpos} 
  a1 categorises the topic as 
       [\textsc{hpos} 0.0–0.5] [\textsc{hpos} 0.0–0.125]
       [\textsc{area} 0.0–0.5] [R 0.0–0.5] [G 0.5–1.0]
  a1 has the words
       puxedu for [G 0.5–1.0] (1.0)
       wawosido for [\textsc{hpos} 0.0–0.5] (0.7)
       pesidimu for [\textsc{area} 0.0–0.5] (0.2)
       wefoto for [R 0.0–0.5] (0.2)
       buwonipo for [G 0.5–1.0] (0.0)
  a1 says: puxedu
  a2 segments the context into 3 objects: 
       h-object-0 h-object-1 h-object-2
  a2 interprets puxedu as
       [G 0.5–1.0] (0.20)
  a2 identifies h-object-0 h-object-1
  a2 says: puxedu?
  a1 points to s-object-0
  a2 categorises the topic as [\textsc{width} 0.5–1.0]
  a2 stores puxedu as [\textsc{width} 0.5–1.0]
\end{verbatim}
